Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-05-30 article/main.tex
écriture section 6

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{tikz}
\usepackage{cellspace}
\usepackage{multirow}
%\usepackage{parskip}
\usetikzlibrary{patterns}

@@

\begin{algorithm}
\begin{algorithmic}[1]
  \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
  \State $v_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$v_0 = S_0[64:64+\ell]$}
  \For{$i=5, \dots, k$}
  \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
  \State $w_i = w_0 + u_i[0:\ell] \bmod 2^{\ell}$ \Comment{$w_i = S_i[0:\ell]$}
  \State $v_i = v_0 + u_i[64:64+\ell] \bmod 2^{\ell}$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
  \State $v'_i = v_i + 1 \bmod 2^{\ell}$
  \State $\mathcal{C} \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
  \If{$\{v_i, v'_i\} \cap \mathcal{C} = \emptyset$}
  \State \Return \textbf{False} \Comment{Bad Guesses}
  \EndIf
  \EndFor
  \State \Return \textbf{True} \Comment{No inconsistency}
  \EndProcedure

@@

  %\State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
  \For{$0 \leq r < 64$} \Comment{Try all rotations}
  \If{$w_i \oplus (X_i \lll r)[0:\ell] \in \left\{v_i, v'_i\right\}$} \Comment{Check eq.~\eqref{eq:find_rotation}}
  \State $\mathcal{H} \gets r \texttt{::} \mathcal{H}$ \Comment{New candidate $r_i$}
  \EndIf
  \EndFor
  \State \Return $\left\{ h\texttt{::} t~|~ h \in \mathcal{H}, t \in \mathcal{T} \right\}$ \Comment{Return $\mathcal{H} \times \mathcal{T}$}
  \EndFunction
  \State
  \Function{ReconstructFullDifference}{...}
  \State \todo{WRITEME}
  \EndFunction
\end{algorithmic}
\caption{Rotations and full difference reconstruction algorithm}
\end{algorithm}

@@

We have implemented the state reconstruction algorithms described above using a
mixture of \textsf{C} (for the computationally expensive parts) and
\textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
to solve CVP instances exactly in dimension 40.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{tikz}
\usepackage{minted}
\usepackage{cellspace}
\usepackage{multirow}
%\usepackage{parskip}
\usetikzlibrary{patterns}

@@

\begin{algorithm}
\begin{algorithmic}[1]
  \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
  \State $v_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$v_0 = S_0[64:64+\ell]$}
  \For{$i=5, \dots, k$}
  \State $u_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{64+\ell}$ \Comment{$u_i = \nabla S_i[0:64+\ell]$}
  \State $w_i = w_0 + u_i[0:\ell] \bmod 2^{\ell}$ \Comment{$w_i = S_i[0:\ell]$}
  \State $v_i = v_0 + u_i[64:64+\ell] \bmod 2^{\ell}$ \Comment{$S_i[64:64+\ell] \in \{v_i, v'_i\}$}
  \State $v'_i = v_i + 1 \bmod 2^{\ell}$
  \State $\mathcal{C}_i \gets \{ w_i \oplus (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$ \Comment{Check eq.~\eqref{eq:consistency}}
  \If{$\{v_i, v'_i\} \cap \mathcal{C}_i = \emptyset$}
  \State \Return \textbf{False} \Comment{Bad Guesses}
  \EndIf
  \EndFor
  \State \Return \textbf{True} \Comment{No inconsistency}
  \EndProcedure

@@

  %\State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
  \For{$0 \leq r < 64$} \Comment{Try all rotations}
  \If{$w_i \oplus (X_i \lll r)[0:\ell] \in \left\{v_i, v'_i\right\}$} \Comment{Check eq.~\eqref{eq:find_rotation}}
  \State $\mathcal{H} \gets r \texttt{::} \mathcal{H}$ \Comment{New candidate $r_i$}
  \EndIf
  \EndFor
  \State \Return $\left\{ h\texttt{::} t~|~ h \in \mathcal{H}, t \in \mathcal{T} \right\}$ \Comment{Return $\mathcal{H} \times \mathcal{T}$}
  \EndFunction
  \State
  \Function{ReconstructFullDifference}{...}
  \State \todo{WRITEME}
  \EndFunction
\end{algorithmic}
\caption{Rotations and full difference reconstruction algorithm}
\label{algo:unknown_2}
\end{algorithm}

@@

We have implemented the state reconstruction algorithms described above using a
mixture of \textsf{C} (for the computationally expensive parts) and
\textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
to solve CVP instances exactly in dimension 42.

In this section, we briefly outline important aspects of our implementations and
present practical results. Our codes are available online at:
\[
  % \url{https://github.com/cbouilla/pcg/}
  \verb|[----REDACTED FOR PEER REVIEW----]|
\]
(and in the supplementary material).

The designer of \textsf{PCG} was kind enough to send us challenge inputs. She
generated random seeds and provided us with the first outputs of the
pseudo-random generator. We were able to reconstruct the seed with an extremely
high confidence level, because they re-generate the same outputs. We emailed
back the seeds and received confirmation that they were indeed correct.

\subsection{Known Increment}

When the increment $c$ is known, algorithm~\ref{algo:known} is all it takes to
reconstruct the internal state of the generator and predict it (or output the
seed). We implemented it in \textsf{C}, using \textsf{OpenMP} to parallelize the
outer loop that guess the least-significant bits of the state. This yields a
simple multi-core implementation. We used the \texttt{gcc 8.3.0} compiler.

We ran it on a server equipped with two 16-core \textsf{Intel Xeon Gold 6130 CPU
  @ 2.10GHz} ``\emph{Skylake}'' CPUs. The inner loop does $2^{37}$ iterations
and terminates in 42.3s, which makes 23 core minutes.

These processors operate at a different frequency depending on the number of
cores used and the type of instructions executed. Our code uses only scalar
instructions, so the CPUs runs at the highest frequency tier when executing
it. Using a single software thread per physical core (each core presents two
hardware execution contexts, commercially called \emph{HyperThreads}) allows the
CPU to run at 2.8Ghz, the maximum ``Turbo'' frequency on all cores. Using one
software thread per hardware thread reduces the frequency to $\approx$ 2.6Ghz,
but allows to better saturate the execution units of the CPU and yields a nearly
20\% speedup overall.

Therefore the algorithm requires $2^{41.67}$ CPU cycles in total; this makes
less than 5 cycles per iteration of the inner loop. We used several
implementation tricks to reach this level of efficiency:

\begin{itemize}
\item We used the \verb|__uint128_t| type provided by most \textsf{C} compilers
  to do 128-bit arithmetic when computing $S_1$ from $S_0$. Apart from that, the
  algorithm has been designed to do mostly 64-bit arithmetic, for the sake of
  efficiency.

\item Looking at the algorithm, it is clear that $U_1$ and $U_2$ are actually
  not needed, so we just don't compute them.

\item $\widetilde{T}_J$ is a function of $w, j$ and $r_J$ (with
  $j=0,1,2$). therefore, for each new value of $w$, we precompute once and for
  all an array indexed by $(J, R_J)$ of the 192 possible values of
  $\widetilde{T}_J$.

\item Pushing the same idea a bit further, we precompute parts of the
  matrix-vector product inside the rounding: this computes a linear combination
  of the rows of $G_3^{-1}$, in which $\widetilde{T}_j$ is the coefficient of
  the $j$-th row. So we precompute the 576 possible products $\widetilde{T}_j \cdot G_3^{-1}[j, k]$.

\item We enumerate the possible rotations in lexicographic order. This means
  that $\widetilde{T}_0$ changes in each iteration while $\widetilde{T}_1$ (resp
  $\widetilde{T}_2$) changes every 64 (resp 4096) iterations. Therefore, in 98\%
  of the iterations, two-thirds of the matrix-vector product inside the rounding
  are the same as from the previous iteration. Therefore, we fully compute the
  matrix-vector product only when $r_1$ changes and only update it when $r_0$
  changes.

\item The rounding operation, when done naively by writing
  $\texttt{llround(x)}$, is actually a bottleneck: it calls a library function
  that accounted for about 20\% of the total running time. We instead used the
  following technique, which correctly returns $\lfloor x \rceil$ whenever
  $|x| < 2^{51}$:
  \begin{minted}{C}
long long fast_round(double x)
{
    union { double d; long long l; } magic; 
    magic.d = x + 6755399441055744.0;
    magic.l <<= 13; 
    magic.l >>= 13;
    return magic.l;
}
\end{minted}
This hack exploits the IEEE754 representation of double-precision floats: the
mantissa lies in bits $[0:52]$ while the sign bit and the exponents take the 12
most significant bits. Adding $2^{52}+2^{51}$ forces the mantissa to shift to
the correct position and inserts an extra 1 bit at position 51. The two shifts
clear the extra bit and the exponent, while correctly expanding the sign bit.
\end{itemize}

\subsection{Unknown Increment}

When the increment $c$ is known, the internal state of \textsf{PCG64} can be
practically reconstructed from $X_0, \dots, X_{42}$ using the algorithms shows
in section~\ref{sec:Cunknown}. Only algorithm~\ref{algo:unknown_1} is
computationally expensive; we implemented it in \textsf{C}, while we implemented
algorithms~\ref{algo:unknown_2} and~\ref{algo:last} in \textsf{Python}.

We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=13$. In
fact, we empirically observed that it returns the right result with probability
$\approx 66\%$ when $\ell=11$ and with probability $99\%$ when
$\ell=12$. Decrementing $\ell$ reduces the workload by a factor of 4. Therefore,
if $T$ denotes the running time when $\ell=13$, we actually used the following
strategy:
\begin{enumerate}
\item Run with $\ell=11$ (if it succeeds, stop).
\item Run with $\ell=11$ with different inputs.
\item Run with $\ell=11$ with different inputs.
\item Run with $\ell=12$.
\item Run with $\ell=13$.
\end{enumerate}
It is not difficult to check that the expected running time of this procedure is
$\frac{1079}{10800} T \approx 0.1 T$. In our implementation, $T = 200,000$ CPU
hours, so the expected running time of the reconstruction procedure is about
$20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
attempt with $\ell=11$ succeeded and took only $12,500$ CPU hours.

The whole process actually ran in 35 wall-clock minutes using 512 cluster nodes,
each equipped with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}
(``\emph{Cascade Lake}''). The actual machine is the
\verb|[----REDACTED FOR PEER REVIEW----]| located at the
\verb|[----ALSO REDACTED----]| computation center. Note that on this particular
parallel computer, running the algorithm with $\ell=13$ would take 10 hours
using the same amount of resources, so the whole procedure \emph{is} practical,
even in the absolute worst case.

The outer loop of algorithm~\ref{algo:unknown_1} makes $2^{2\ell - 1}$
iterations while the inner loop makes $2^{30}$ iterations. Using a single
hardware execution context, we measured that one of the outer loop takes between
41.5s and 44s (apparently not all nodes of the cluster are running at exactly
the same speed, potentially because of ``turbo boost'' adjustments and thermal
constraints; most of the time, they take exactly... 42 seconds). Because of this
variability, we implemented a master-slave work distribution pattern, in which a
master process dispatches iterations of the outer loop to slave processes. This
also made checkpointing very easy. We used \textsf{MPI} for inter-process
communication.

With $\ell=11$, the whole process took $2^{56.74}$ CPU cycles, which makes less
than 6 cycles per iteration of the inner loop. We used essentially the same
implementation tricks discussed above. However, this time we had to additionally
implement the \textsc{ConsistencyCheck} procedure, which is called in the inner
loop. We observed that the set of possible candidate values $\mathcal{C}$ only
depends on $w_0$ (the variable of the outer loop). Therefore, before entering
the inner loop, we precompute a bit field of size $2^\ell$ describing
$\mathcal{C}_i$. To simplify the implementation, we flatten them by computing
$\mathcal{C} = \cup_i \mathcal{C}_i$. This slightly increase the probability of
false positives.
